Revisiting~the~Classics: Online~RL~in~the~Programmable~Dataplane

### 2021-05-21 paper/paper.tex
Shorter tables (and stop making Windows cry)

?? We are the first to bring fully online RL to the dataplane!
\end{abstract}

@@

@@

We would be remiss to believe that P4 and the PSA define all capabilities that programmable hardware supports.
Many compatible devices have legacies long predating these developments. 
For instance, many-core SOC-based Netronome SmartNIC~\parencite{netronome-smartnic}, and NetFPGA SUME~\parencite{DBLP:journals/micro/ZilbermanACM14} NICs allow virtually arbitrary programs to be specified and executed.
The first of these compiles from P4$\rightarrow$Micro-C$\rightarrow$bytecode (with Micro-C externs), while the latter can combine the P4$\rightarrow$NetFPGA toolchain~\parencite{DBLP:conf/fpga/IbanezBMZ19} with arbitrary circuit externs---both exposing further device capabilities beyond the specification.
This pattern extends to other SmartNICs, such as NVIDIA BlueField~\parencite{nvidia-bluefield} and Xilinx Alveo~\parencite{xilinx-alveo} devices.
Even Intel Tofino~\parencite{barefoot-intel} ASICs, which architecturally mirror the PSA, expose additional matching capabilities and ALU functions via the Tofino Native Architecture.
Regardless, floating-point operations key to ML/RL workloads are very rarely supported outside of deep learning accelerators (which, in turn, lack packet switching functionality).
Moreover, the limited memory/block RAM and per-packet timing constraints endemic to these devices make this class of in-NIC offloading challenging.

\begin{insight}
	Low-port density programmable network devices often have spare resources and extra capabilities beyond the PSA specification which can aid asynchronous, local compute.
\end{insight}

@@

				\draw[dotted, ->, bend left = 30] (state) -- (agent);
				\draw[->] (agent) -- (action);
				\draw[dotted, ->] (action) -- (state);
		\end{tikzpicture}}
		\caption{Theory: state measurement, action computation, and learning are zero-cost.}
	\end{subfigure}
	\begin{subfigure}{0.45\linewidth}
		\resizebox{0.75\linewidth}{!}{
			\centering
			\begin{tikzpicture}
				\node[circle, draw] (state) {$S$};
				\node[circle, draw] (state') at ($(state) + (0, -1.5)$) {$S'$};
				\node[circle, draw] (state'') at ($(state') + (0, -1.5)$) {$S''$};

@@

The solution we propose is to make use of the recent wave of programmable network devices to \emph{bring reinforcement learning to the data plane}---placing state measurement ($t_1$), low-cost decision-making processes ($t_2$), and controlled systems ($t_3$) as close to one another as possible.
In networks, actions are most likely to be installed on backbone switches, bump-in-the-wire NICs or middleboxes, and in the NICs of end-hosts.
Ideally, these functions which comprise an RL agent would all be collocated on the same chip or device, but this is easier said than done. 
Both programmable devices and the network environment make this more difficult, as we'll examine in the sequel.

@@

We then return again to what classical RL methods can offer us; in particular, the simple linear function approximation given by \emph{tile coding} \cite[pp.\ \numrange{217}{221}]{RL2E}.
The idea is simple: a policy is represented by sets of tiles, each covering one or more dimensions of the input state with several overlapping tilings (offset stepwise to provide generalisation).
A state vector produces a single `hit' in each tiling, which provides both the set of action preference lists to sum during inference \emph{and} the set of tiles to update at the next timestep.
Crucially, there are no data dependencies between calculating any tile hits for an input state vector.
This has important benefits for parallelisation.
To select an action, we can produce a task for each tiling---retrieving the list of action values contributed by the tile activated by the input state.
An aggregate step sums up all action preference lists, and the action with the highest value is then selected.
Updating the policy (with a new $\delta_t$ value) has no aggregate step, and as tile indices map to disjoint memory regions no locks are required.

@@

\section{Design and Implementation}\label{sec:design}
Based on the design principles, problems, and potential solutions outlined throughout \cref{sec:motivation}, we present our design for an in-NIC, task-independent, online reinforcement learning system---\emph{\approachshort{} (\approach)}.
At a high level, \approachshort{} is designed to use the auxiliary compute exposed by general SmartNIC devices to offer low-latency online learning, scaling according to available on-chip resources at build time.

@@

\approachshort{} runs on one more cores of a SmartNIC to convert fixed-point state measurements from the environment into a stream of actions using a stored policy.
As an example, this might be to map flow state and performance measurements into a queue priority for future packets from that flow, or to compute and apply a rate limit to preserve quality of service.
These dedicated cores are then responsible for processing requests, computing actions, and updating the underlying policy in real time.
Combined with reward measurements, this policy can then be updated or trained from scratch entirely on the NIC, acting as a fully online RL agent.
An input state vector \emph{always} induces an action, and if desired updates the policy using either an included reward or one retrieved from memory according to a (configurable) key placed alongside the state.
This allows for simultaneous control and learning over independent systems by the same agent (i.e., optimising several flows with their own reward measures).

@@

@@

@@

\subsection{Intra-Agent Communication}
Even with embarrassingly parallel problems, optimising for latency requires meticulous care in how work is passed out and aggregated.
This is truer still when moving from the moderately fine-grained, efficient control of classical RL methods ($\sim$\SI{2}{\milli\second}) to its logical limit (tens of \si{\micro\second}).
Ordinarily, the serialisation/marshalling of requests, responses, and safe shared data access can incur significant extra overheads.
On-chip execution and the nature of action preference computation allow us to use lockless atomic aggregation, removing the overheads of explicit messaging/packetisation.
Moreover, physically adjacent functional units may have special-purpose shared registers or share a small fast cache to accelerate communication further.

@@

We use a simple first-fit work placement algorithm run on the NFP, placing the largest work item into the least loaded thread of the least loaded core.
The cost of any individual work item (according to its dimension count and memory location) was estimated offlin.
This work allocation and any cached per-work-item values are recomputed when any configuration is installed or changed.

@@

@@

@@

@@

@@

\begin{table}
	\caption{Latencies and computation times for \approachshort{} versus commodity hardware hosts. On-device execution is crucial in not only lowering latencies, but in reducing tail latencies. Lower is better, with the best marked \emph{in bold}.\label{tab:lats}}
	\resizebox{\linewidth}{!}{
		\expandableinput ../tables/build/latency
	}
\end{table}

\begin{table}
	\caption{Action and update throughputs for \approachshort{} versus commodity hardware hosts. Most designs cannot scale online performance with additional cores. Higher is better, with the best marked \emph{in bold}.\label{tab:tputs}}
	\resizebox{\linewidth}{!}{
		\expandableinput ../tables/build/tput
	}
\end{table}

@@

@@

>>>>

?? We are the first to bring fully online RL to the dataplane!

?? Emphasise here and elsewhere: scales with core count, but with some designs (\Indfw) achieve better latency and per-core throughput with just one core/worker!
\end{abstract}

@@

?? TODO: trim down. Too ``thesis-y'', rather than ``conference-y''.
?? Save this elsewhere 'til then?

@@

We would be remiss to believe that P4 and the PSA define all capabilities that programmable hardware supports.
Many compatible devices have legacies long predating these developments. 
For instance, many-core SOC-based Netronome SmartNIC~\parencite{netronome-smartnic}, and NetFPGA SUME~\parencite{DBLP:journals/micro/ZilbermanACM14} NICs allow virtually arbitrary programs to be specified and executed.
The first of these compiles from P4$\rightarrow$Micro-C$\rightarrow$bytecode (with Micro-C externs), while the latter can combine the P4$\rightarrow$NetFPGA toolchain~\parencite{DBLP:conf/fpga/IbanezBMZ19} with arbitrary circuit externs---both exposing further device capabilities beyond the specification.
This pattern extends to other SmartNICs, such as NVIDIA BlueField~\parencite{nvidia-bluefield} and Xilinx Alveo~\parencite{xilinx-alveo} devices.

Currently, low port density devices like the above are most likely to have specific general-purpose compute, as they are designed to suit high-performance offloading and middlebox development.
Even Intel Tofino~\parencite{barefoot-intel} ASICs, which architecturally mirror the PSA, expose additional matching capabilities and \emph{arithmetic logic unit} (ALU) functions via the Tofino Native Architecture.
Regardless, floating-point operations key to ML/RL workloads are very rarely supported outside of deep learning accelerators (which, in turn, lack packet switching functionality).
Moreover, the limited memory/block RAM and per-packet timing constraints endemic to these devices make this class of in-NIC offloading challenging.

\begin{insight}
	Current low-port density programmable network devices often have spare resources and extra capabilities beyond the PSA specification which can aid asynchronous, local compute.
\end{insight}

@@

				\draw[dotted, ->, bend left = 30] (state) -- (agent);
				\draw[->] (agent) -- (action);
				\draw[dotted, ->] (action) -- (state);
		\end{tikzpicture}}
		\caption{Theory: state measurement, action computation, and learning are zero-cost.}
	\end{subfigure}
	\hspace{0.05\linewidth}
	\begin{subfigure}{0.45\linewidth}
		\resizebox{0.75\linewidth}{!}{
			\centering
			\begin{tikzpicture}
				\node[circle, draw] (state) {$S$};
				\node[circle, draw] (state') at ($(state) + (0, -1.5)$) {$S'$};
				\node[circle, draw] (state'') at ($(state') + (0, -1.5)$) {$S''$};

@@

The solution we propose is to make use of the recent wave of programmable network devices to \emph{bring reinforcement learning to the data plane}---referring again to \cref{fig:state-slip}, we would place place state measurement ($t_1$), low-cost decision-making processes ($t_2$), and controlled systems ($t_3$) as close to one another as possible.
In networks, actions are most likely to be installed on backbone switches, bump-in-the-wire NICs or middleboxes, and in the NICs of end-hosts.
Ideally, these functions which comprise an RL agent would all be collocated on the same chip or device, but this is easier said than done. 
Both programmable devices and the network environment make this more difficult, as we'll examine in the sequel.

@@

We then return again to what classical RL methods can offer us; in particular, the simple linear function approximation given by \emph{tile coding} \cite[pp.\ \numrange{217}{221}]{RL2E}.
The idea is simple: a policy is represented by sets of tiles, each covering one or more dimensions of the input state with several overlapping tilings (offset stepwise to provide generalisation).
A state vector produces a single `hit' in each tiling, which provides both the set of action preference lists to sum during inference \emph{and} the set of tiles to update at the next timestep.
Crucially, this internal representation ensures that there are no data dependencies between any tile hit calculations for an input state vector.
This has important benefits for parallelisation.
To select an action, we can produce a task for each tiling---retrieving the list of action values contributed by the tile activated by the input state.
An aggregate step sums up all action preference lists, and the action with the highest value is then selected.
Updating the policy (with a new $\delta_t$ value) has no aggregate step, and as tile indices map to disjoint memory regions no locks are required.

@@

\section{Design and Implementation}\label{sec:design}
Based on the design principles, problems, and potential solutions outlined throughout \cref{sec:motivation}, we present our design for an in-NIC, task-independent, online reinforcement learning system---\emph{\approachshort{} (\approach)}.
At a high level, \approachshort{} is designed to use the auxiliary compute exposed by general SmartNIC devices to offer low-latency online learning, scaling according to available on-chip resources at build time.
As the allocation of cores/chip area is set ahead of time by a framework or system administrator, \approachshort{} (\Coopfw) agents enumerate themselves at runtime initialisation.

@@

\approachshort{} runs on one or more cores of a SmartNIC to convert fixed-point state measurements from the environment into a stream of actions using a stored policy.
As an example, this might be to map flow state and performance measurements into a queue priority for future packets from that flow, or to compute and apply a rate limit to preserve quality of service.
These dedicated cores are then responsible for processing requests, computing actions, and updating the underlying policy in real time.
Combined with reward measurements, this policy can then be updated or trained from scratch entirely on the NIC, acting as a fully online RL agent.
An input state vector \emph{always} induces an action, and if desired updates the policy using either an included reward or one retrieved from memory according to a (configurable) key placed alongside the state.
This allows for simultaneous control and learning over independent systems by the same agent (i.e., optimising several flows with their own reward measures).

@@

?? Move a bunch of this to the NFP basics chapter.

@@

?? While we \emph{could} use bitshifts in tile coding, we don't due to the strict limits it places upon policy design.

@@

\subsection{Intra-Agent Communication}
Even with embarrassingly parallel problems, optimising for latency requires meticulous care in how work is passed out and aggregated.
This is truer still when moving from the moderately fine-grained, efficient control of classical RL methods ($\sim$\SI{1}{\milli\second}) to its logical limit (tens of \si{\micro\second}).
Ordinarily, the serialisation/marshalling of requests, responses, and safe shared data access can incur significant extra overheads.
On-chip execution and the nature of action preference computation allow us to use lockless atomic aggregation, removing the overheads of explicit messaging/packetisation.
Moreover, physically adjacent functional units may have special-purpose shared registers or share a small fast cache to accelerate communication further.

?? Move much of the below to the NFP basics section?

@@

We use a simple first-fit work placement algorithm run on the NFP, placing the largest work item into the least loaded thread of the least loaded core.
The cost of any individual work item (according to its dimension count and memory location) was estimated offline.
This work allocation and any cached per-work-item values are recomputed when any configuration is installed or changed.
Naturally, for $n$ tasks and $m$ threads this is $\mathcal{O}{\left(n\log{m}\right)}$: two find/update min operations into binary heaps per task, storing $m/8$ and $\le8$ costs respectively.

@@

?? Need to mention that this is a theoretical improvement; the results are basically opposite to this.
?? Main benefit is then, I guess, the redux in storage requirements?

@@

?? Achieving very tight latency bounds on \Coopfw{} requires quite a decent core allocation. (i.e., need at least 11 workers to make a gain in online t'put, 5 cores needed to win in throughput).

@@

?? Will these look better \emph{after} the results?

@@

?? Any way to move some of this sooner? I.e., this is ``why move DDoS prevention here'', Stefanos suggested backporting some of this to justify ``why \emph{RL} in-NIC''?

@@

\begin{table}
	\caption{Latencies and computation times for \approachshort{} versus commodity hardware hosts. On-device execution is crucial in not only lowering latencies, but in reducing tail latencies. Lower is better, with the best marked \emph{in bold}.\label{tab:lats}}
	\resizebox{\linewidth}{!}{
		\expandableinput ../tables/build/latency-only32
	}
\end{table}

\begin{table}
	\caption{Action and update throughputs for \approachshort{} versus commodity hardware hosts. Most designs cannot scale online performance with additional cores. Higher is better, with the best marked \emph{in bold}.\label{tab:tputs}}
	\resizebox{\linewidth}{!}{
		\expandableinput ../tables/build/tput-only32
	}
\end{table}

@@

?? Urgently describe and explain why lower bitdepth gives *worse* metrics.
?? Mainly, register width.
?? We also do stuff to batch read as many action vals as possible!! And yet this happens!!
?? Also explain that Integer math does worse due to it falling back to BLAS in numpy (i.e., a float optimiser).
?? We might have good int results if I do it in Rust?

@@

?? Future directions for community: want to see more general-purpose async compute on-chip! Larger form-factors!

?? Are there other ``old ways of doing things'' worth looking at?
